Reanalysis Step 7: Sample Comparison Table

### 2025-09-04 reanalysis-7-missingness_patterns.qmd
data comparison

---
title: "Reanalysis Step 7: Missingness Pattern Analysis"
subtitle: "Comparing characteristics of excluded vs included patients"
author: "Erik Westlund"
date: "`r Sys.Date()`"
format: html
---

@@

```{r setup}
#| include: false
# Setup analysis environment
source("dependencies.R")
source("functions.R")
setup_analysis(seed = 2025)

# Additional packages for this analysis
library(knitr)
library(tableone)
library(ggplot2)
library(gridExtra)

# Initialize logging
log_file <- init_log("missingness_pattern_analysis")

# Load the original data before exclusions
# Use the load_and_prepare_data function to get processed variables
original_data <- load_and_prepare_data(reanalysis_data_file_path)

# Load imputed datasets (this is a list of 20 imputed dataframes)
imputed_datasets <- readRDS(file.path(reanalysis_data_dir, "imputed_datasets.rds"))

log_message("Data loaded successfully")
```

## Motivation

When we performed multiple imputation, the observed treatment effect increased. This suggests that those with missing data (excluded from complete case analysis) may differ systematically from those included.

This analysis compares baseline characteristics between:

- **Excluded**: Patients with missing outcome data (excluded from complete case analysis)
- **Included**: Patients with complete outcome data (included in analysis)

## Identify Excluded vs Included Patients

```{r identify-groups}
# Use cohort_id if it exists to identify excluded vs included
if ("cohort_id" %in% names(original_data)) {
  # cohort_id has values "excluded" and "submitted"
  # "excluded" = patients with missing outcomes who were excluded from complete case analysis
  # "submitted" = patients with complete outcomes who were included
  original_data$excluded <- original_data$cohort_id == "excluded"
  log_message("Using cohort_id to identify excluded patients")
  log_message(paste("Cohort distribution:", 
                   sum(original_data$cohort_id == "excluded"), "excluded,",
                   sum(original_data$cohort_id == "submitted"), "submitted"))
} else if ("outcome_va_vi_binary" %in% names(original_data)) {
  # Use missing outcome as the exclusion criterion
  original_data$excluded <- is.na(original_data$outcome_va_vi_binary)
  log_message("Using missing outcome_va_vi_binary to identify excluded patients")
} else if ("outcome_va_logMAR" %in% names(original_data)) {
  # Use missing outcome VA
  original_data$excluded <- is.na(original_data$outcome_va_logMAR)
  log_message("Using missing outcome_va_logMAR to identify excluded patients")
} else {
  stop("Cannot identify excluded patients - no cohort_id or outcome variables found")
}

# Verify the excluded variable has two levels
if (length(unique(original_data$excluded)) < 2) {
  log_message(paste("Warning: All patients are", 
                   ifelse(all(original_data$excluded), "excluded", "included")))
  log_message("Check data processing - there should be both excluded and included patients")
}

# Summary counts
n_total <- nrow(original_data)
n_included <- sum(!original_data$excluded)  # FALSE = included
n_excluded <- sum(original_data$excluded)   # TRUE = excluded
pct_excluded <- round(100 * n_excluded / n_total, 1)

log_message(sprintf("Total patients (eyes): %d", n_total))
log_message(sprintf("Included (submitted/complete outcome): %d (%.1f%%)", n_included, 100 - pct_excluded))
log_message(sprintf("Excluded (missing outcome): %d (%.1f%%)", n_excluded, pct_excluded))

# Create summary table
summary_df <- data.frame(
  Group = c("Included (submitted)", "Excluded", "Total"),
  N = c(n_included, n_excluded, n_total),
  Percentage = c(round(100 - pct_excluded, 1), pct_excluded, 100)
)
```

```{r summary-table}
#| echo: false
kable(summary_df,
      caption = "Sample Size Summary",
      col.names = c("Group", "N (eyes)", "Percentage (%)"),
      align = c("l", "r", "r"))
```

## Variable Selection

```{r variable-selection}
# Get variables used in outcome and propensity score models
# Log what columns exist in the processed data
log_message(paste("Total columns in processed data:", length(names(original_data))))
log_message(paste("Sample columns:", paste(names(original_data)[1:min(10, length(names(original_data)))], collapse = ", ")))

# Variables from the actual matching and outcome formulas in functions.R
# Demographics created by load_and_prepare_data
demographic_vars <- c("age", "age_cat", "gender_cat", "race_ethnic_cat", "insurance_cat")

# Comorbidity scores (these ARE in the raw data)
comorbidity_vars <- c("CCI", "DCSI")

# Vision and treatment variables
vision_vars <- c("baseline_VA_logMAR")

# Treatment flags (these ARE in the raw data)
treatment_vars <- c("other_inject", "anti_VEGF", "focal_laser_flag", "PRP_flag")

# Other eye conditions
eye_condition_vars <- c("glaucoma_bef_hitplus_cat", "otherretina_bef_hitplus_cat", 
                        "catsurg_before_hitplus_cat")

# DR severity (created by the function)
dr_vars <- c("dr_severity")

# Treatment/exposure (created by the function)
exposure_var <- "ever_lapse_binary"

# Outcome variables (created by the function)
outcome_vars <- c("outcome_va_logMAR", "outcome_va_vi_binary")

# Combine all variables
all_vars <- c(demographic_vars, comorbidity_vars, vision_vars, treatment_vars,
              eye_condition_vars, dr_vars, exposure_var, outcome_vars)

# Check which variables exist in the data
available_vars <- intersect(all_vars, names(original_data))
missing_vars <- setdiff(all_vars, names(original_data))

if (length(missing_vars) > 0) {
  log_message(paste("Variables not found in data:", paste(missing_vars, collapse = ", ")))
}

log_message(paste("Analyzing", length(available_vars), "variables"))
```

## Descriptive Statistics by Exclusion Status

```{r descriptive-stats}
# Create Table 1 comparing excluded vs included
# Specify which variables should be treated as factors
factor_vars <- c("age_cat", "gender_cat", "race_ethnic_cat", "insurance_cat",
                 "glaucoma_bef_hitplus_cat", "otherretina_bef_hitplus_cat", 
                 "catsurg_before_hitplus_cat", "dr_severity", "ever_lapse_binary")

# Ensure factor variables are factors
for (var in factor_vars) {
  if (var %in% names(original_data)) {
    if (!is.factor(original_data[[var]])) {
      original_data[[var]] <- as.factor(original_data[[var]])
    }
  }
}

# Create Table 1
table1 <- CreateTableOne(
  vars = available_vars,
  strata = "excluded",
  data = original_data,
  factorVars = intersect(factor_vars, available_vars),
  includeNA = TRUE,
  test = TRUE,
  smd = TRUE
)

# Print table with standardized mean differences
# Label the strata clearly
table1_print <- print(table1, 
                      smd = TRUE, 
                      showAllLevels = TRUE,
                      formatOptions = list(big.mark = ","),
                      printToggle = FALSE)

# Update column names for clarity
colnames(table1_print)[colnames(table1_print) == "FALSE"] <- "Included (n)"
colnames(table1_print)[colnames(table1_print) == "TRUE"] <- "Excluded (n)"

kable(table1_print, 
      caption = "Comparison of Baseline Characteristics: Excluded vs Included Patients")
```

## Key Differences Visualization

```{r key-differences}
#| fig-height: 10
#| fig-width: 12

# Extract SMDs with more detail for factor variables
# Get the table itself to access all SMD information
table1_matrix <- print(table1, smd = TRUE, printToggle = FALSE)

# Extract SMDs manually from the table for each row
# This will give us SMDs for each level of factor variables
smd_data <- data.frame()

# Process each variable in the table
for (var in available_vars) {
  if (var %in% names(original_data)) {
    # Check if it's a factor or character that should be treated as factor
    is_factor_var <- (var %in% factor_vars) || is.factor(original_data[[var]]) || is.character(original_data[[var]])

    if (is_factor_var) {
      # For factor/categorical variables, calculate SMD for each level
      # Convert to factor if needed
      if (!is.factor(original_data[[var]])) {
        original_data[[var]] <- as.factor(original_data[[var]])
      }

      var_levels <- levels(original_data[[var]])
      for (level in var_levels) {
        # Calculate proportion in each group
        prop_excluded <- mean(original_data[[var]][original_data$excluded] == level, na.rm = TRUE)
        prop_included <- mean(original_data[[var]][!original_data$excluded] == level, na.rm = TRUE)

        # Calculate SMD for this level
        # SMD for proportions = (p1 - p2) / sqrt((p1*(1-p1) + p2*(1-p2))/2)
        pooled_var <- (prop_excluded * (1 - prop_excluded) + prop_included * (1 - prop_included)) / 2
        smd_val <- if (pooled_var > 0) {
          abs(prop_excluded - prop_included) / sqrt(pooled_var)
        } else {
          0
        }

        smd_data <- rbind(smd_data, data.frame(
          Variable = paste0(var, ": ", level),
          SMD = smd_val,
          VarType = "Factor",
          BaseVar = var,
          stringsAsFactors = FALSE
        ))
      }
    } else if (is.numeric(original_data[[var]])) {
      # For continuous variables, calculate single SMD
      if (sum(!is.na(original_data[[var]])) > 0) {
        # Calculate means
        mean_excluded <- mean(original_data[[var]][original_data$excluded], na.rm = TRUE)
        mean_included <- mean(original_data[[var]][!original_data$excluded], na.rm = TRUE)

        # Calculate pooled standard deviation
        var_excluded <- var(original_data[[var]][original_data$excluded], na.rm = TRUE)
        var_included <- var(original_data[[var]][!original_data$excluded], na.rm = TRUE)

        # Check if we have valid variances
        if (!is.na(var_excluded) && !is.na(var_included)) {
          sd_pooled <- sqrt((var_excluded + var_included) / 2)

          smd_val <- if (!is.na(sd_pooled) && sd_pooled > 0) {
            abs(mean_excluded - mean_included) / sd_pooled
          } else {
            0
          }
        } else {
          smd_val <- 0
        }

        smd_data <- rbind(smd_data, data.frame(
          Variable = var,
          SMD = smd_val,
          VarType = "Continuous",
          BaseVar = var,
          stringsAsFactors = FALSE
        ))
      }
    }
  }
}

# Remove NA values and order by absolute SMD
smd_data <- smd_data[!is.na(smd_data$SMD), ]
smd_data <- smd_data[order(abs(smd_data$SMD), decreasing = TRUE), ]

# Get top differences (for display purposes, limit to top 20)
smd_data_top <- head(smd_data, 20)

# Identify variables with substantial differences (SMD > 0.1)
substantial_diff <- smd_data[abs(smd_data$SMD) > 0.1, ]

# Plot 1: SMD plot (showing top 20 differences)
p1 <- ggplot(smd_data_top, aes(x = reorder(Variable, abs(SMD)), y = SMD)) +
  geom_point(aes(color = VarType), size = 3) +
  geom_point(data = smd_data_top[abs(smd_data_top$SMD) > 0.1, ],
             color = "red", size = 3.5) +
  geom_hline(yintercept = c(-0.1, 0, 0.1), linetype = c(2, 1, 2), alpha = 0.5) +
  coord_flip() +
  labs(title = "Standardized Mean Differences: Excluded vs Included (Top 20)",
       subtitle = "Red overlay indicates SMD > 0.1; Factor levels shown separately",
       x = "Variable",
       y = "Standardized Mean Difference",
       color = "Variable Type") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))

print(p1)
```

## Focus on Key Variables

```{r key-variables}
# Analyze key variables of interest
key_results <- data.frame(
  Variable = character(),
  Included = character(),
  Excluded = character(),
  P_value = numeric(),
  SMD = numeric(),
  stringsAsFactors = FALSE
)

# Helper function to format results
format_continuous <- function(data, var, excluded_flag) {
  subset_data <- data[data$excluded == excluded_flag & !is.na(data[[var]]), var]
  sprintf("%.2f (%.2f)", mean(subset_data, na.rm = TRUE), sd(subset_data, na.rm = TRUE))
}

format_binary <- function(data, var, excluded_flag, level = "1") {
  subset_data <- data[data$excluded == excluded_flag, ]
  n <- sum(!is.na(subset_data[[var]]))
  n_level <- sum(subset_data[[var]] == level, na.rm = TRUE)
  sprintf("%d (%.1f%%)", n_level, 100 * n_level / n)
}

# Analyze lapsing rate
if ("ever_lapse_binary" %in% available_vars) {
  lapse_included <- mean(original_data$ever_lapse_binary[!original_data$excluded] == 1, na.rm = TRUE)
  lapse_excluded <- mean(original_data$ever_lapse_binary[original_data$excluded] == 1, na.rm = TRUE)
  lapse_p <- chisq.test(table(original_data$excluded, original_data$ever_lapse_binary))$p.value

  key_results <- rbind(key_results, data.frame(
    Variable = "Lapsed from care",
    Included = sprintf("%.1f%%", 100 * lapse_included),
    Excluded = sprintf("%.1f%%", 100 * lapse_excluded),
    P_value = lapse_p,
    SMD = abs(lapse_excluded - lapse_included) / sqrt((lapse_included * (1 - lapse_included) + 
                                                        lapse_excluded * (1 - lapse_excluded)) / 2)
  ))
}

# Analyze baseline VA
if ("baseline_VA_logMAR" %in% available_vars) {
  va_included <- mean(original_data$baseline_VA_logMAR[!original_data$excluded], na.rm = TRUE)
  va_excluded <- mean(original_data$baseline_VA_logMAR[original_data$excluded], na.rm = TRUE)
  va_p <- t.test(original_data$baseline_VA_logMAR ~ original_data$excluded)$p.value

  key_results <- rbind(key_results, data.frame(
    Variable = "Baseline VA (logMAR)",
    Included = sprintf("%.3f", va_included),
    Excluded = sprintf("%.3f", va_excluded),
    P_value = va_p,
    SMD = abs(va_excluded - va_included) / sd(original_data$baseline_VA_logMAR, na.rm = TRUE)
  ))
}

# Analyze CCI
if ("CCI" %in% available_vars) {
  cci_included <- mean(original_data$CCI[!original_data$excluded], na.rm = TRUE)
  cci_excluded <- mean(original_data$CCI[original_data$excluded], na.rm = TRUE)
  cci_p <- t.test(original_data$CCI ~ original_data$excluded)$p.value

  key_results <- rbind(key_results, data.frame(
    Variable = "Charlson Comorbidity Index",
    Included = sprintf("%.2f", cci_included),
    Excluded = sprintf("%.2f", cci_excluded),
    P_value = cci_p,
    SMD = abs(cci_excluded - cci_included) / sd(original_data$CCI, na.rm = TRUE)
  ))
}

# Format and display
key_results$P_value <- format.pval(key_results$P_value, digits = 3)
key_results$SMD <- round(key_results$SMD, 3)

kable(key_results,
      caption = "Key Variable Comparisons",
      col.names = c("Variable", "Included", "Excluded", "P-value", "SMD"))
```

## Missingness Patterns by DR Severity

```{r missingness-by-dr}
#| fig-height: 6
#| fig-width: 10

if ("dr_severity" %in% available_vars) {
  # Calculate missingness rate by DR severity
  miss_by_dr <- original_data %>%
    group_by(dr_severity) %>%
    summarise(
      n_total = n(),
      n_excluded = sum(excluded),
      pct_excluded = 100 * n_excluded / n_total,
      .groups = 'drop'
    )

  # Plot missingness by DR severity
  p2 <- ggplot(miss_by_dr, aes(x = dr_severity, y = pct_excluded)) +
    geom_col(fill = "steelblue", alpha = 0.7) +
    geom_text(aes(label = sprintf("%.1f%%", pct_excluded)), 
              vjust = -0.5, size = 4) +
    labs(title = "Outcome Missingness by DR Severity",
         subtitle = "Percentage of patients excluded due to missing outcomes",
         x = "DR Severity",
         y = "% Excluded") +
    theme_minimal() +
    theme(plot.title = element_text(size = 14, face = "bold"))

  print(p2)

  # Table of missingness by DR severity
  kable(miss_by_dr,
        caption = "Missingness Patterns by DR Severity",
        col.names = c("DR Severity", "Total N", "N Excluded", "% Excluded"),
        digits = 1)
}
```

## Missingness Patterns by Treatment and Lapse Status

```{r missingness-by-treatment}
#| fig-height: 8
#| fig-width: 12

# Create 2x2 plots for treatment and lapse patterns
plots <- list()

# Plot 1: Missingness by treatment status
if ("any_treatment" %in% available_vars) {
  miss_by_treatment <- original_data %>%
    group_by(any_treatment) %>%
    summarise(
      n_total = n(),
      n_excluded = sum(excluded),
      pct_excluded = 100 * n_excluded / n_total,
      .groups = 'drop'
    )

  plots[[1]] <- ggplot(miss_by_treatment, aes(x = any_treatment, y = pct_excluded)) +
    geom_col(fill = "coral", alpha = 0.7) +
    geom_text(aes(label = sprintf("%.1f%%", pct_excluded)), 
              vjust = -0.5, size = 4) +
    labs(title = "Missingness by Treatment Status",
         x = "Any Treatment", y = "% Excluded") +
    theme_minimal()
}

# Plot 2: Missingness by lapse status
if ("ever_lapse_binary" %in% available_vars) {
  miss_by_lapse <- original_data %>%
    mutate(lapse_status = factor(ever_lapse_binary, 
                                 levels = c(0, 1),
                                 labels = c("Not Lapsed", "Lapsed"))) %>%
    group_by(lapse_status) %>%
    summarise(
      n_total = n(),
      n_excluded = sum(excluded),
      pct_excluded = 100 * n_excluded / n_total,
      .groups = 'drop'
    )

  plots[[2]] <- ggplot(miss_by_lapse, aes(x = lapse_status, y = pct_excluded)) +
    geom_col(fill = "darkgreen", alpha = 0.7) +
    geom_text(aes(label = sprintf("%.1f%%", pct_excluded)), 
              vjust = -0.5, size = 4) +
    labs(title = "Missingness by Lapse Status",
         x = "Lapse Status", y = "% Excluded") +
    theme_minimal()
}

# Plot 3: Joint distribution
if ("any_treatment" %in% available_vars && "ever_lapse_binary" %in% available_vars) {
  miss_joint <- original_data %>%
    mutate(lapse_status = factor(ever_lapse_binary, 
                                 levels = c(0, 1),
                                 labels = c("Not Lapsed", "Lapsed"))) %>%
    group_by(any_treatment, lapse_status) %>%
    summarise(
      n_total = n(),
      n_excluded = sum(excluded),
      pct_excluded = 100 * n_excluded / n_total,
      .groups = 'drop'
    )

  plots[[3]] <- ggplot(miss_joint, aes(x = any_treatment, y = pct_excluded, 
                                       fill = lapse_status)) +
    geom_col(position = position_dodge(width = 0.8), alpha = 0.7) +
    labs(title = "Missingness by Treatment and Lapse Status",
         x = "Treatment Status", y = "% Excluded",
         fill = "Lapse Status") +
    theme_minimal() +
    scale_fill_manual(values = c("Not Lapsed" = "blue", "Lapsed" = "red"))
}

# Arrange plots
if (length(plots) > 0) {
  # Remove NULL elements from plots list
  plots <- plots[!sapply(plots, is.null)]

  if (length(plots) == 1) {
    print(plots[[1]])
  } else if (length(plots) > 1) {
    do.call(grid.arrange, c(plots, list(ncol = 2)))
  }
}
```

## Comparison Across Imputed Datasets

```{r imputation-comparison}
# For key variables, compare means across imputed datasets for originally missing vs complete
# This shows how imputation filled in the missing values

imputation_summary <- data.frame()

# Get IDs of included patients (those not excluded)
id_var <- if("cohort_id" %in% names(original_data)) "cohort_id" else "e_mrn_deidentified"
included_data <- original_data[!original_data$excluded, ]
included_ids <- unique(included_data[[id_var]])

for (var in c("baseline_VA_logMAR", "CCI", "DCSI")) {
  if (var %in% names(original_data)) {  # Check if variable exists in original data
    # Get which observations were originally missing
    originally_missing <- is.na(original_data[[var]])

    # Calculate mean across imputations for originally missing
    if (sum(originally_missing, na.rm = TRUE) > 0) {
      means_missing <- sapply(imputed_datasets, function(df) {
        # Match IDs to find corresponding rows in imputed data
        imputed_ids <- df[[id_var]]
        matching_rows <- imputed_ids %in% included_ids
        missing_and_included <- originally_missing[match(imputed_ids, original_data[[id_var]])] & matching_rows

        if (sum(missing_and_included, na.rm = TRUE) > 0) {
          mean(df[[var]][missing_and_included], na.rm = TRUE)
        } else {
          NA
        }
      })

      # Calculate mean for originally complete
      means_complete <- sapply(imputed_datasets, function(df) {
        imputed_ids <- df[[id_var]]
        matching_rows <- imputed_ids %in% included_ids
        complete_and_included <- !originally_missing[match(imputed_ids, original_data[[id_var]])] & matching_rows

        if (sum(complete_and_included, na.rm = TRUE) > 0) {
          mean(df[[var]][complete_and_included], na.rm = TRUE)
        } else {
          NA
        }
      })

      # Remove NAs before calculating statistics
      means_missing <- means_missing[!is.na(means_missing)]
      means_complete <- means_complete[!is.na(means_complete)]

      if (length(means_missing) > 0 && length(means_complete) > 0) {
        imputation_summary <- rbind(imputation_summary, data.frame(
          Variable = var,
          Originally_Complete = sprintf("%.3f (%.3f)", 
                                       mean(means_complete), sd(means_complete)),
          Imputed_for_Missing = sprintf("%.3f (%.3f)", 
                                        mean(means_missing), sd(means_missing)),
          Difference = sprintf("%.3f", mean(means_missing) - mean(means_complete))
        ))
      }
    }
  }
}

if (nrow(imputation_summary) > 0) {
  kable(imputation_summary,
        caption = "Comparison of Imputed Values: Originally Missing vs Complete",
        col.names = c("Variable", "Originally Complete", "Imputed (for Missing)", "Difference"))
}
```

## Summary and Implications

```{r summary}
# Summarize key findings
significant_differences <- substantial_diff$Variable

# Calculate effect size measures
lapse_or_excluded_vs_included <- NA
if ("ever_lapse_binary" %in% available_vars) {
  # Calculate odds ratio of being excluded for lapsed vs not lapsed
  lapse_table <- table(Lapse = original_data$ever_lapse_binary, 
                      Excluded = original_data$excluded)
  if (ncol(lapse_table) == 2 && nrow(lapse_table) == 2) {
    # OR = (lapsed & excluded * not lapsed & not excluded) / 
    #      (lapsed & not excluded * not lapsed & excluded)
    lapse_or_excluded_vs_included <- (lapse_table["1", "TRUE"] * lapse_table["0", "FALSE"]) / 
                                     (lapse_table["1", "FALSE"] * lapse_table["0", "TRUE"])
  }
}

log_message("Analysis completed successfully")
```

### Key Findings

**Variables with Substantial Differences (SMD > 0.1):**
```{r}
#| echo: false
if (length(significant_differences) > 0) {
  for (var in significant_differences) {
    cat("- ", var, "\n")
  }
} else {
  cat("No variables showed substantial differences (SMD > 0.1)\n")
}
```

```{r lapse-missingness-summary}
#| echo: false
# Create summary of lapse status and missingness relationship
if (!is.na(lapse_or_excluded_vs_included)) {
  lapse_summary <- data.frame(
    Measure = c("Odds ratio of exclusion (lapsed vs not lapsed)",
                "Interpretation"),
    Value = c(sprintf("%.2f", lapse_or_excluded_vs_included),
              ifelse(lapse_or_excluded_vs_included > 1.5,
                     "Lapsed patients were substantially more likely to have missing outcomes",
                     "No substantial difference in missingness by lapse status"))
  )
  kable(lapse_summary, 
        caption = "Lapse Status and Missingness Relationship",
        col.names = c("Measure", "Value"))
} else {
  cat("Lapse status and missingness relationship could not be calculated.\n")
}
```

```{r finalize-log}
#| include: false
# Finalize logging
log_message("Missingness pattern analysis completed successfully")
final_log <- finalize_log(success = TRUE)
```

>>>>

---
title: "Reanalysis Step 7: Sample Comparison Table"
subtitle: "Comparing Original, Complete Cases, and Imputed Data"
author: "Erik Westlund"
date: "`r Sys.Date()`"
format: html
---

@@

```{r setup}
#| include: false
source("dependencies.R")
source("functions.R")
source("settings.R")
setup_analysis(seed = 2025)

library(tableone)
library(knitr)
library(dplyr)

# Initialize logging
log_file <- init_log("sample_comparison")
```

## Load Data from Three Sources

```{r load-data}
# 1. Load original data (from analysis.Rmd)
data_path <- file.path(s_root, "Gina's Project/df_Final_240909.csv")
original_data <- load_and_prepare_data(data_path)
original_complete <- apply_exclusions(original_data)  # Remove missing CCI/DCSI

log_message(paste0("Original data after exclusions: ", nrow(original_complete), " observations"))

# 2. Load complete cases from reanalysis (reanalysis-1)
complete_cases_file <- file.path(reanalysis_data_dir, "df_complete_original.rds")
if (!file.exists(complete_cases_file)) {
  stop("Complete cases data not found. Run reanalysis-1-data_preparation.qmd first.")
}
complete_cases_data <- readRDS(complete_cases_file)

log_message(paste0("Complete cases data: ", nrow(complete_cases_data), " observations"))

# 3. Load imputed datasets (reanalysis-2)
imputed_file <- file.path(reanalysis_data_dir, "imputed_datasets.rds")
if (!file.exists(imputed_file)) {
  stop("Imputed datasets not found. Run reanalysis-2-multiple_imputation.qmd first.")
}
imputed_datasets <- readRDS(imputed_file)
n_imputations <- length(imputed_datasets)

log_message(paste0("Loaded ", n_imputations, " imputed datasets"))
```

## Create Comparison Table

```{r prepare-variables}
# Define variables for Table 1
vars_continuous <- c("age", "baseline_VA_logMAR", "CCI", "DCSI")
vars_categorical <- c("gender_cat", "race_ethnic_cat", "insurance_cat", "age_cat",
                      "dr_severity", "ever_lapse_binary", 
                      "glaucoma_bef_hitplus_cat", "otherretina_bef_hitplus_cat",
                      "catsurg_before_hitplus_cat",
                      "anti_VEGF", "PRP_flag", "focal_laser_flag", "other_inject",
                      "outcome_va_vi_binary")

all_vars <- c(vars_continuous, vars_categorical)

# Function to compute summary statistics for continuous variables
summarize_continuous <- function(data, var) {
  values <- data[[var]]
  values <- values[!is.na(values)]
  if (length(values) == 0) {
    return(data.frame(
      variable = var,
      mean = NA,
      sd = NA,
      median = NA,
      q25 = NA,
      q75 = NA,
      n_missing = nrow(data),
      pct_missing = 100
    ))
  }

  data.frame(
    variable = var,
    mean = mean(values, na.rm = TRUE),
    sd = sd(values, na.rm = TRUE),
    median = median(values, na.rm = TRUE),
    q25 = quantile(values, 0.25, na.rm = TRUE),
    q75 = quantile(values, 0.75, na.rm = TRUE),
    n_missing = sum(is.na(data[[var]])),
    pct_missing = 100 * sum(is.na(data[[var]])) / nrow(data)
  )
}

# Function to compute summary statistics for categorical variables
summarize_categorical <- function(data, var) {
  tbl <- table(data[[var]], useNA = "ifany")
  props <- prop.table(table(data[[var]], useNA = "no"))

  levels_df <- data.frame(
    variable = paste0(var, "_", names(props)),
    n = as.vector(table(data[[var]], useNA = "no")),
    pct = 100 * as.vector(props)
  )

  # Add missing info
  n_missing <- sum(is.na(data[[var]]))
  levels_df$n_missing <- n_missing
  levels_df$pct_missing <- 100 * n_missing / nrow(data)

  levels_df
}

# Process each dataset
process_dataset <- function(data, name) {
  # Continuous variables
  cont_summaries <- do.call(rbind, lapply(vars_continuous, function(v) {
    if (v %in% names(data)) {
      summarize_continuous(data, v)
    } else {
      data.frame(variable = v, mean = NA, sd = NA, median = NA, 
                 q25 = NA, q75 = NA, n_missing = NA, pct_missing = NA)
    }
  }))

  # Categorical variables
  cat_summaries <- do.call(rbind, lapply(vars_categorical, function(v) {
    if (v %in% names(data)) {
      summarize_categorical(data, v)
    } else {
      data.frame(variable = v, n = NA, pct = NA, n_missing = NA, pct_missing = NA)
    }
  }))

  list(continuous = cont_summaries, categorical = cat_summaries, dataset = name)
}
```

```{r create-table}
# Process original data
original_summary <- process_dataset(original_complete, "Original")

# Process complete cases
complete_summary <- process_dataset(complete_cases_data, "Complete Cases")

# Process imputed data (average across imputations)
# For imputed data, we need to average across all imputations
imputed_summaries <- lapply(1:n_imputations, function(i) {
  process_dataset(imputed_datasets[[i]], paste0("Imputation_", i))
})

# Average the imputed results
average_imputed_continuous <- do.call(rbind, lapply(vars_continuous, function(v) {
  # Extract values for this variable across all imputations
  imp_values <- lapply(imputed_summaries, function(s) {
    s$continuous[s$continuous$variable == v, ]
  })

  # Average the statistics
  if (nrow(imp_values[[1]]) > 0) {
    data.frame(
      variable = v,
      mean = mean(sapply(imp_values, function(x) x$mean), na.rm = TRUE),
      sd = mean(sapply(imp_values, function(x) x$sd), na.rm = TRUE),
      median = mean(sapply(imp_values, function(x) x$median), na.rm = TRUE),
      q25 = mean(sapply(imp_values, function(x) x$q25), na.rm = TRUE),
      q75 = mean(sapply(imp_values, function(x) x$q75), na.rm = TRUE),
      n_missing = 0,  # No missing in imputed data
      pct_missing = 0
    )
  } else {
    data.frame(variable = v, mean = NA, sd = NA, median = NA, 
               q25 = NA, q75 = NA, n_missing = NA, pct_missing = NA)
  }
}))

# For categorical variables, average the percentages
all_cat_vars <- unique(unlist(lapply(imputed_summaries, function(s) s$categorical$variable)))
average_imputed_categorical <- do.call(rbind, lapply(all_cat_vars, function(v) {
  # Extract values for this variable level across all imputations
  imp_values <- lapply(imputed_summaries, function(s) {
    s$categorical[s$categorical$variable == v, ]
  })

  # Remove empty results
  imp_values <- imp_values[sapply(imp_values, nrow) > 0]

  if (length(imp_values) > 0) {
    data.frame(
      variable = v,
      n = mean(sapply(imp_values, function(x) x$n), na.rm = TRUE),
      pct = mean(sapply(imp_values, function(x) x$pct), na.rm = TRUE),
      n_missing = 0,
      pct_missing = 0
    )
  } else {
    data.frame(variable = v, n = NA, pct = NA, n_missing = NA, pct_missing = NA)
  }
}))

imputed_summary <- list(
  continuous = average_imputed_continuous,
  categorical = average_imputed_categorical,
  dataset = "Imputed (Average)"
)
```

```{r display-table}
# Create formatted table for continuous variables
format_continuous_table <- function(original, complete, imputed) {
  vars <- unique(c(original$variable, complete$variable, imputed$variable))

  table_df <- data.frame(
    Variable = vars,
    Original = NA,
    Complete_Cases = NA,
    Imputed = NA,
    stringsAsFactors = FALSE
  )

  for (i in 1:length(vars)) {
    v <- vars[i]

    # Original
    orig_row <- original[original$variable == v, ]
    if (nrow(orig_row) > 0 && !is.na(orig_row$mean)) {
      table_df$Original[i] <- sprintf("%.2f (%.2f)", orig_row$mean, orig_row$sd)
    }

    # Complete cases
    comp_row <- complete[complete$variable == v, ]
    if (nrow(comp_row) > 0 && !is.na(comp_row$mean)) {
      table_df$Complete_Cases[i] <- sprintf("%.2f (%.2f)", comp_row$mean, comp_row$sd)
    }

    # Imputed
    imp_row <- imputed[imputed$variable == v, ]
    if (nrow(imp_row) > 0 && !is.na(imp_row$mean)) {
      table_df$Imputed[i] <- sprintf("%.2f (%.2f)", imp_row$mean, imp_row$sd)
    }
  }

  table_df
}

# Create formatted table for categorical variables
format_categorical_table <- function(original, complete, imputed) {
  vars <- unique(c(original$variable, complete$variable, imputed$variable))

  table_df <- data.frame(
    Variable = vars,
    Original = NA,
    Complete_Cases = NA, 
    Imputed = NA,
    stringsAsFactors = FALSE
  )

  for (i in 1:length(vars)) {
    v <- vars[i]

    # Original
    orig_row <- original[original$variable == v, ]
    if (nrow(orig_row) > 0 && !is.na(orig_row$n)) {
      table_df$Original[i] <- sprintf("%.0f (%.1f%%)", orig_row$n, orig_row$pct)
    }

    # Complete cases
    comp_row <- complete[complete$variable == v, ]
    if (nrow(comp_row) > 0 && !is.na(comp_row$n)) {
      table_df$Complete_Cases[i] <- sprintf("%.0f (%.1f%%)", comp_row$n, comp_row$pct)
    }

    # Imputed
    imp_row <- imputed[imputed$variable == v, ]
    if (nrow(imp_row) > 0 && !is.na(imp_row$n)) {
      table_df$Imputed[i] <- sprintf("%.1f (%.1f%%)", imp_row$n, imp_row$pct)
    }
  }

  table_df
}

# Create the tables
continuous_table <- format_continuous_table(
  original_summary$continuous,
  complete_summary$continuous,
  imputed_summary$continuous
)

categorical_table <- format_categorical_table(
  original_summary$categorical,
  complete_summary$categorical,
  imputed_summary$categorical
)

# Display continuous variables
kable(continuous_table, 
      caption = "Continuous Variables: Mean (SD)",
      col.names = c("Variable", "Original Data", "Complete Cases", "Imputed (Average)"))

# Display categorical variables
kable(categorical_table,
      caption = "Categorical Variables: N (%)",
      col.names = c("Variable", "Original Data", "Complete Cases", "Imputed (Average)"))

# Sample sizes
sample_sizes <- data.frame(
  Dataset = c("Original (after exclusions)", "Complete Cases", "Imputed (each)"),
  N = c(nrow(original_complete), nrow(complete_cases_data), nrow(imputed_datasets[[1]]))
)

kable(sample_sizes, caption = "Sample Sizes")
```

```{r finalize}
# Finalize log
finalize_log()
```
